---
title: "Fourier Regression in NIMBLE"
author: "Adam B. Smith"
date: "`r Sys.Date()`"
output:
  word_document: default
    toc: true
  html_document: default
    toc: true
---

```{r setup, include=FALSE}`r ''`
knitr::opts_chunk$set(
	fig.path = "man/figures/",
	echo = TRUE
)
```

Fourier regression uses a set of sine waves to approximate an unknown function. Since sine waves are cyclic, Fourier regression is best used for data that are cyclic (e.g., temperature across a series of days). However, any smooth function can be approximated to an arbitrary degree using a set of sine waves, though they will likely extrapolate wildly outside the range of the data.  Here, I will demonstrate Fourier regression of non-cyclic data in `nimble`.

## Basic Fourier regression

```{r, quiet = TRUE}
library(AICcmodavg) # used for function AICc()
library(bayesplot) # for graphing
library(coda) # for model diagnostics
library(boot) # used for inverse logit function
library(nimble) # the workhorse
```

Create some fake data:
```{r}
beta0 <- -0.4
beta1 <- 0.6
beta2 <- -1.4
beta3 <- 4.7
sigma <- 0.9

n <- 1000 # sample size
x <- runif(n, -1, 1) # predictor variable
x <- sort(x) # sort to help in plotting later... no effect on model
y <- beta0 + beta1 * x + beta2 * x^2 + beta3 * x^3 + rnorm(n, 0, sigma) +
	ifelse(x > 0.5, -17, 0)

plot(x, y)
```

Our fake data is very non-smooth!

Fourier regression uses a seris of sine waves with different wavelengths (frequencies). In the models below, we'll use 20 sine waves. Let's see what the look like.

```{r}
min_x <- min(x)
max_x <- max(x)
range_x <- max_x - min_x

n_harmonics <- 20
harmonics <- 1:n_harmonics

sines <- cosines <- matrix(NA_real_, nrow = n, ncol = n_harmonics)

for (i in 1:n_harmonics) {
	sines[ , i] <- sin((2 * pi * harmonics[i] * x) / range_x)
	cosines[ , i] <- cos((2 * pi * harmonics[i] * x) / range_x)
}

colnames(sines) <- paste0('sine_', harmonics)
colnames(cosines) <- paste0('cosine_', harmonics)

par(mfrow = c(2, 4))
wl <- c(1, 2, 3, 4, 8, 15, 20)
for (i in wl) {
	plot(x, sines[ , i], col = 'red', pch = 16, type = 'l', main = harmonics[i])
	lines(x, cosines[ , i], col = 'blue', pch = 16, type = 'l', main = harmonics[i])
}
```

Of the 20 sine waves we'll be using, these are the ones with of 1, 2, 3, 4, 8, 15, and 20 wavelengths across the range of the data (which spans (-1, 1)).

## Frequentist Fourier regression
Now, let's use a frequentist approach to approximate the data. We will compare a Fourier regression to a model that has nearly the same form used to generate the data (the "pseudogenerative" model). This pseudogenerative model does not include the abrupt break the data has, but it will be close. The Fourier regression uses 20 sine waves.

```{r}
# setup for Fourier model
sine_cosine_names <- c(colnames(sines), colnames(cosines))
xy_matrix <- cbind(y, x)

data <- cbind(xy_matrix, sines, cosines)
data <- as.data.frame(data)

# Fourier model
form <- paste0('y ~ 1 + ', paste(sine_cosine_names, collapse = ' + '))
lm_fourier <- lm(form, data = data)

# linear model that approximates the process we used to generate the data
lm_pseudogenerative <- lm(y ~ 1 + x + I(x^2) + I(x^3), data = data)
```

Let's see how the models compare:
```{r}
summary(lm_pseudogenerative)
summary(lm_fourier)

data <- data[order(data$x), ]

pred_pseudogenerative <- predict(lm_pseudogenerative, data)
pred_fourier <- predict(lm_fourier, data)

par(mfrow = c(1, 2))
plot(x, y, main = 'Pseudogenerative model')
lines(x, pred_pseudogenerative, col = 'red')

plot(x, y, main = 'Fourier model')
lines(x, pred_fourier, col = 'red')
```

The Fourier models does better at capturing the overall trend, though the smaller sine waves also impart bias in the fine-scale structure.

## Redo, but with more waves
Let's see how adding sine waves increases the fit (and overfit) of the model. We'll explore models with 1 to `r 2^7` sine waves, and identify teh most parsimonious model using AICc.
```{r}
n_harmonics <- 2^(0:7)
aicc <- numeric()
par(mfrow = c(2, 4))

for (n_harmonic in n_harmonics) {

	harmonic <- 1:n_harmonic
	n_waves <- length(harmonic)

	sines <- cosines <- matrix(NA_real_, nrow = n, ncol = n_waves)

	for (i in 1:n_waves) {
		sines[ , i] <- sin((2 * pi * harmonic[i] * x) / range_x)
		cosines[ , i] <- cos((2 * pi * harmonic[i] * x) / range_x)
	}
	colnames(sines) <- paste0('sine_', harmonic)
	colnames(cosines) <- paste0('cosine_', harmonic)

	sine_cosine_names <- c(colnames(sines), colnames(cosines))
	form <- paste0('y ~ 1 + ', paste(sine_cosine_names, collapse = ' + '))
	xy_matrix <- cbind(y, x)

	data <- cbind(xy_matrix, sines, cosines)
	data <- as.data.frame(data)
	data <- data[order(data$x), ] # helps with plotting

	lm_fourier <- lm(form, data = data)

	pred_sines <- predict(lm_fourier, data)

	plot(x, y, main = paste('Number of waves: ', n_harmonic))
	lines(x, pred_sines, col = 'red')

	aicc <- c(aicc, AICc(lm_fourier))

} 
```

```{r}
min_aicc <- which.min(aicc)
aiccs <- data.frame(n_harmonic = n_harmonics, AICc = aicc)

plot(aiccs)
points(aiccs$n_harmonic[min_aicc], aiccs$AIC[min_aicc], pch = 16, col = 'red')
```


### NIMBLE: NEEDS COSINES ADDED!!!
```{r}
n_waves <- 10
sines <- matrix(NA_real_, nrow = n, ncol = n_waves)

for (i in 1:n_waves) {
	sines[ , i] <- sin(x_rescale * freq[i])
}
colnames(sines) <- paste0('freq_', freq[1:n_waves] / 2)

ones <- matrix(1, nrow = n, ncol = 1)
x_matrix <- cbind(ones, sines)

n_betas <- ncol(x_matrix)

data <- list(
	y = y
)

constants <- list(
	n = n,
	n_betas = n_betas,
	x_matrix = x_matrix
)

inits <- list(
	beta_hat = rep(0, n_betas),
	tau_hat = 1
)

code <- nimbleCode({

	# priors
	for (i in 1:n_betas) {
		beta_hat[i] ~ dnorm(0, tau = 0.01)
	}

	# likelihood
	for (i in 1:n) {
		y[i] ~ dnorm(mu_hat[i], tau = tau_hat)
		mu_hat[i] <- inprod(beta_hat[1:n_betas], x_matrix[i, 1:n_betas])

	}

})

# black magic
model <- nimbleModel(
	code = code, # our model
	constants = constants, # constants
	data = data, # data
	inits = inits, # initialization values
	check = TRUE, # any errors?
	calculate = FALSE
)

model$initializeInfo()
model$calculate()

# configure
monitors <- c('beta_hat', 'tau_hat')
conf <- configureMCMC(
	model,
	monitors = monitors,
	print = TRUE,
	enableWAIC = FALSE # Watanabe's AIC... useful, but takes time
)

build <- buildMCMC(conf)

# compile
compiled <- compileNimble(model, build, showCompilerOutput = FALSE)

# MCMC
chains <- runMCMC(
	compiled$build,
	niter = 6000, # iterations
	nburnin = 1000, # burn-in
	thin = 1, # thinning rate
	nchains = 4, # number of chains
	inits = inits, # initialization values
	progressBar = TRUE,
	samplesAsCodaMCMC = TRUE, # for using coda package for plots
	summary = TRUE, # calculate summaries across chains
	WAIC = FALSE, # Watanabe's AIC... useful, but takes time
	perChainWAIC = FALSE
)

# gather mean of coefficient estimates
beta_hats <- rep(NA_real_, n_betas)
for (i in 1:n) {
	beta_hat <- paste0('beta_hat[', i, ']')
	beta_hats[i] <- chains$summary$all.chains[beta_hat, 'Mean']
}

# plot
plot(x, y)
y_hat <- x_matrix %*% matrix(beta_hats, ncol = 1)
lines(x, y_hat, col = 'red')
legend('bottomright', legend = c('data', 'prediction'), pch = c(1, NA), lwd = c(NA, 1), col = c('black', 'red'))
```

# Combining Fourier regression with non-trigonometric variables
```{r}
beta0 <- -0.4
beta1 <- 0.6
beta2 <- 2.4
beta3 <- 4.7
sigma <- 0.9

n <- 200 # sample size
x <- runif(n, -1, 1) # predictor variable
x <- sort(x) # sort to help in plotting later... no effect on model
y <- beta0 + beta1 * x + beta2 * x^2 + beta3 * x^3 + rnorm(n, 0, sigma)

plot(x, y)
```

```{r}
min_x <- min(x)
max_x <- max(x)
range_x <- max_x - min_x

n_harmonics <- 20
harmonics <- 1:n_harmonics

sines <- cosines <- matrix(NA_real_, nrow = n, ncol = n_harmonics)

for (i in 1:n_harmonics) {
	sines[ , i] <- sin((2 * pi * harmonics[i] * x) / range_x)
	cosines[ , i] <- cos((2 * pi * harmonics[i] * x) / range_x)
}

colnames(sines) <- paste0('sine_', harmonics)
colnames(cosines) <- paste0('cosine_', harmonics)

xy_matrix <- cbind(y, x)

data <- cbind(xy_matrix, sines, cosines)
data <- as.data.frame(data)
```

```{r}
form_generative <- 'x + I(x^2) + I(x^3)'
form_generative_missing <- 'x + I(x^2)'
form_sines <- paste(colnames(sines), collapse = ' + ')

form_generative_sines <- paste0('y ~ 1 + ', form_generative, ' + ', form_sines)
form_generative_missing_sines <- paste0('y ~ 1 + ', form_generative_missing, ' + ', form_sines)
form_generative <- paste0('y ~ 1 + ', form_generative)
form_generative_missing <- paste0('y ~ 1 + ', form_generative_missing)
form_sines <- paste0('y ~ 1 + ', form_sines)

lm_generative <- lm(as.formula(form_generative), data = data)
lm_generative_missing <- lm(as.formula(form_generative_missing), data = data)
lm_sines <- lm(as.formula(form_sines), data = data)
lm_generative_fourier <- lm(as.formula(form_generative_sines), data = data)
lm_generative_missing_fourier <- lm(as.formula(form_generative_missing_sines), data = data)
```

```{r}
model_coeffs <- c('(Intercept)', 'x', 'I(x^2)', 'I(x^3)')
coeffs <- matrix(c(
		c(beta0, beta1, beta2, beta3),
		coefficients(lm_generative)[model_coeffs],
		coefficients(lm_generative_missing)[model_coeffs],
		coefficients(lm_generative_fourier)[model_coeffs],
		coefficients(lm_generative_missing_fourier)[model_coeffs]
	), byrow = TRUE, ncol = 4
)

colnames(coeffs) <- model_coeffs
rownames(coeffs) <- c('real', 'generative', 'generative missing x^3', 'generative fourier', 'generative missing x^3 with fourier')
coeffs
```

Note that that for the models combining `x` and fourier analysis, the coefficients for the `x` terms are very different from the "truth", sometimes by a lot! Let's try to improve this by choosing orthogonal sine and cosine components.

```{r}
n_harmonics <- 20
harmonics <- 1:n_harmonics

x_grad <- seq(min_x, max_x, length.out = 10000)

sines <- cosines <- matrix(NA_real_, nrow = 10000, ncol = n_harmonics)

for (i in 1:n_harmonics) {
	sines[ , i] <- sin((2 * pi * harmonics[i] * x_grad) / range_x)
	cosines[ , i] <- cos((2 * pi * harmonics[i] * x_grad) / range_x)
}

colnames(sines) <- paste0('sine_', harmonics)
colnames(cosines) <- paste0('cosine_', harmonics)

integral <- data.frame()
for (i in harmonics) {
	for (j in harmonics) {
		integral[i, j] <- sum(sines[ , paste0('sine_', i)] * cosines[ , paste0('cosine_', i)])
	}
}
colnames(integral) <- colnames(cosines)
rownames(integral) <- colnames(sines)


```

